---
title: "US sex specific fetal- and infant mortality. Data quality and relationships"
author: "Jonas Schöley"
date: "August 18, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, fig.keep = TRUE)
```

This short report looks into possible relationships between fetal- and infant mortality rates, sex differences thereof and the secondary sex ratio.

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

load("../../priv/data/pre-harmonized/us_fdeath_1989-2010.RData")
load("../../priv/data/pre-harmonized/us_ideath_1989-2010.RData")
```

Data preparation
----------------

```{r}
# merge data on births, fetal- and infant deaths
fideath <- bind_rows(infant = bind_rows(ideath),
                     fetus = bind_rows(fdeath),
                     .id = "type")
rm(fdeath, ideath)

# prepare 
fideath %>%
  mutate(
    date_of_delivery_ym =
      ymd(paste(date_of_delivery_y, date_of_delivery_m, "01", sep = "-")),
    date_of_conception_ym =
      round_date(date_of_delivery_ym - weeks(gestation_at_delivery_w), unit = "month"),
    date_of_conception_y =
      year(date_of_conception_ym),
    death = # death indicator (always TRUE for data from the fetal death file)
      !is.na(age_at_death_d) | type == "fetus"
  ) -> fideath

# summarise
glimpse(fideath)
summary(fideath)

# pattern of missing values in sex and gestation at delivery for fetus and infants
fideath %>% group_by(type, date_of_delivery_y, death) %>%
  summarise(gestation_at_delivery_na = sum(is.na(gestation_at_delivery_w)) / n(),
            sex_na = sum(is.na(sex)) / n())

# pattern of missing values in sex over gestation for fetal deaths
fideath %>% filter(type == "fetus") %>%
  group_by(gestation_at_delivery_w) %>%
  summarise(sex_na = sum(is.na(sex)) / n())
```

The sex of the dead fetus is not known in about 50% of the cases throughout the years. However, the NAs are solely concentrated prior to a gestation of 20 weeks and sex is always known thereafter, except for cases in which the gestational age is undetermined. For infants the sex is always known. From 1995 until 2002 the gestation at delivery is unknown for about 15% of the infants (no significant difference between those who died and those who survived). In the other periods it is known in more than 99% of the cases. I am not sure about the reason for this pattern.

We exclude fetal deaths prior to week 20 from the analysis. We exclude cases where gestation at delivery or sex is unknown.

```{r}
# subset to gestation > 20 weeks,
# remove cases where sex or gestation at delivery is unknown
fideath %>%
  filter(gestation_at_delivery_w >= 20,
         !is.na(sex), !is.na(gestation_at_delivery_w)) -> fideath_sub
```

The number of registered fetal deaths varies throughout time with 5 periods clearly segmented by different levels: 1989-1992, 1992-1995, 1995-2001, 2001-2007, 2007-2011. Excluding cases with unknown sex, gestation at delivery and gestation at delivery less than 20 weeks corrects for the segmentation. Time series of births and infant deaths show no segmentation (analysis not shown).

```{r}
# fetal deaths over time complete data
fideath %>%
  filter(type == "fetus") %>%
  count(date_of_delivery_ym) %>%
  ggplot(aes(x = date_of_delivery_ym, y = n)) +
  geom_line() +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "month") +
  theme(aspect.ratio = 0.07)

# fetal deaths over time for cases with known sex and gestation
fideath_sub %>%
  filter(type == "fetus") %>%
  count(date_of_delivery_ym) %>%
  ggplot(aes(x = date_of_delivery_ym, y = n)) +
  geom_line() +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "month") +
  theme(aspect.ratio = 0.07)
```

Infant- and fetal period mortality
----------------------------------

What are the infant- and fetal mortality rates by sex and period? Following the "National Center of Health Statistics" in the US, we calculate fetal mortality as the number of fetal deaths per fetal deaths and births in period $p$; and infant mortality as the number of infant deaths per numer of births in period $p$. Having no information on the number of births or infant deaths in years 1992, 1993, and 1994 we exclude these periods from the calculation.

We can't find any obvious relationship between the levels of fetal- and infant mortality in a given period.

```{r}
# infant and fetal period mortality rates
fideath_sub %>%
  filter(!date_of_delivery_y %in% 1992:1994) %>%
  group_by(date_of_delivery_y, date_of_delivery_ym, sex) %>%
  summarise(
    fetalmort  = sum(type == "fetus") / n(),
    infantmort = sum(type == "infant" & death == TRUE) / sum(type == "infant")
  ) -> fimort_period

# infant versus fetal period mortality rates by sex and period
ggplot(fimort_period, aes(x = fetalmort, y = infantmort, colour = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "blue", aes(group = sex)) +
  geom_smooth(method = "lm", group = 1, se = FALSE, colour = "black") +
  facet_wrap(~date_of_delivery_y, scales = "fixed") +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Fetal mortality shows some unusual characteristics. It increased between 1991 and 1995, declining thereafter with a steep decline between 2003 and 2004. The reason for the steep decline is unclear. Infant mortality decreases in a regular fashion.

```{r}
# fetal and infant mortality over time by sex
ggplot(fimort_period,
       aes(x = date_of_delivery_ym)) +
  geom_line(aes(y = fetalmort)) +
  geom_line(aes(y = infantmort), alpha = 0.3) +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "month") +
  facet_wrap(~sex) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

No obvious relationship between fetal- and infant mortality sex-ratio in a given period can be found.

```{r}
# fetal and infant mortality sex ratio by period
fimort_period %>%
  gather(... = fetalmort:infantmort) %>%
  unite(var, sex, key) %>%
  spread(var, value) %>%
  mutate(sex_ratio_fmort = Male_fetalmort / Female_fetalmort,
         sex_ratio_imort = Male_infantmort / Female_infantmort) -> fimort_period_sex_ratio

# plot fetal and infant mortality sex ratio by period
ggplot(fimort_period_sex_ratio, aes(x = sex_ratio_fmort, y = sex_ratio_imort)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~date_of_delivery_y) +
  coord_equal()
```

Infant- and fetal conception cohort mortality
---------------------------------------------

We repeat the same analysis, but this time with mortality rates calculated for monthly conception cohorts. The cohort of conception is defined as the month of delivery minus the months of gestation at delivery. Each conception cohort $c$ includes individuals from delivery cohort $p$ and $p+1$. Therefore we leave out conception cohorts which are incomplete due to missing information on births/deaths in the following years.

In opposition to the period perspective, we see a clear correlation between the levels of fetal and infant mortality when comparing the rates on a conception cohort basis. This makes sense as the conception cohorts concept captures similar historical conditions during development while the period concept only captures similar historical conditionsl during death.

```{r}
# infant and fetal conception cohort mortality rates
fideath_sub %>%
  filter(!date_of_conception_y %in% c(1988, 1991:1994, 2002:2004, 2010)) %>%
  group_by(date_of_conception_y, date_of_conception_ym, sex) %>%
  summarise(
    fetalmort  = sum(type == "fetus") / n(),
    infantmort = sum(type == "infant" & death == TRUE) / sum(type == "infant")
  ) -> fimort_cohort

# infant versus fetal period mortality rates by sex and conception cohort
ggplot(fimort_cohort, aes(x = fetalmort, y = infantmort, colour = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "blue", aes(group = sex)) +
  geom_smooth(method = "lm", group = 1, se = FALSE, colour = "black") +
  facet_wrap(~date_of_conception_y, scales = "fixed") +
  coord_fixed()
```

No obvious relationship between fetal- and infant mortality sex-ratio in a given conception cohort can be found.

```{r}
# fetal and infant mortality sex ratio by period
fimort_cohort %>%
  gather(... = fetalmort:infantmort) %>%
  unite(var, sex, key) %>%
  spread(var, value) %>%
  mutate(sex_ratio_fmort = Male_fetalmort / Female_fetalmort,
         sex_ratio_imort = Male_infantmort / Female_infantmort) -> fimort_cohort_sex_ratio

# plot fetal and infant mortality sex ratio by period
ggplot(fimort_cohort_sex_ratio,
       aes(x = sex_ratio_fmort, y = sex_ratio_imort)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_equal() +
  facet_wrap(~date_of_conception_y)
```
